On probabilistic one-vs-one classification

### 2025-04-02 paper.tex
Minor changes

\newtheorem{prop}{Proposition}
\newtheorem{thm}{Theorem}

@@

\usepackage{lastpage}
\jmlrheading{23}{2022}{1-\pageref{LastPage}}{1/21; Revised 5/22}{9/22}{21-0000}{Author One and Author Two}

@@

\author{\name Ondrej Šuch \email ondrejs@savbb.sk \\
       \addr Matematický ústav SAV\\
       Ďumbierska 1\\
       Banská Bystrica, 974 01, Slovakia
       \AND
       \name Peter Novotný \email two@cs.berkeley.edu \\
       \addr Division of Computer Science\\
       University of California\\
       Berkeley, CA 94720-1776, USA}

@@

An exact solution to this problem is provided by the Bayes classifier. To simplify the discussion we will assume that $p_i(x) > 0$ for all $x$ and $P_i > 0$ for all $i$. Commonly used classifiers such as linear discriminant analysis, (penalized) logistic regression or support vector machines.

@@

\begin{align}
	p_{ij}^\Pi(x) = \frac{p_i(x) P_1}{p_i(x)P_1 + p_j(x)P_2}.
\end{align}

Knowledge of vector $\boldsymbol{p}^\Pi$ is  equivalent to knowing the functions $p^\Pi_{ij}(x)$. Indeed, if $p_{i}> 0$ for all $x$, then we have
\begin{align}
p_{ij}^\Pi(x) = \frac{p_i^\Pi(x)}{p_i^\Pi(x) + p_j^\Pi(x)}. \label{eq:bt1}
\end{align}

@@

In practice one does not have know true probability distributions and hence cannot construct Bayes classifier. However, given a sample from the distribution, we can train binary classifiers to provide approximations $r_{ij}(x)$ to pairwise Bayes classifiers $p_{ij}(x)$. The goal of probabilistic one-vs-one classification is to deduce for any $x$ in the feature space $X$ an estimate $\hat{\boldsymbol{p}}(x)$ for $\boldsymbol{p}^\Pi(x)$. One does it by finding approximate solution of the system of equations
%\begin{align}
%
%\end{align}
%
%
%Suppose that $K$ classes $C_1, \ldots, C_K$ are distributed according to probability distributions $p_i$ on a  space $X$. Let us denote the $r_{ij}(x)$ the output of Bayes (binary) classifier for $x$ in $X$ yielding the probability $p(x \in C_i \mid x\in C_i \cup C_j )$. Then we have
%\begin{equation}
%r_{ij}(x)= \frac {p_i(x)}{p_i(x) + p_j(x)}.
%\end{equation}
%
%
%The system of equations is called Bradley--Terry equations.  The equations in \eqref{eq:bt1}  can be transformed to a system of linear equations. The theoretical basis for one-vs-one classification framework is provided by the following result.
%
%
%
%Suppose we have estimators $\hat r_{ij}$ for Bayes predictions $r_{ij}$ for all pairs $i\not= j$. A \emph{coupling method} obtains a multi-class probabilistic estimate $\boldsymbol{\hat p}= (\hat p_1, \ldots, \hat p_K)$ that satisfy
\begin{equation}
	{r}_{ij}(x) \approx \frac {\hat p_i(x)}{\hat p_i(x) + \hat p_j(x)}. \label{eq:bt2}
\end{equation}

@@

Equations forming system \eqref{eq:bt2} are called Bradley-Terry equations.
The system is usually inconsistent and additional assumptions are needed. A method to solve Bradley-Terry equations is called a coupling method. Its input can be represented by a matrix with off-diagonal entries being $r_{ij}(x)$.

Let us outline what kind of conditions may be desirable in a coupling method. In Section \ref{sec:des:exact} we will describe exact mathematical criteria, while in \ref{sec:des:inexact} we shall describe more flexible properties.

@@

@@

\emph{Symmetry} refers to the action of the permutation group which acts both on the classes as well as on predictions. The action of a permutation $\sigma$ extends to the action on pairwise matrix by 
\begin{equation}
	\hat r^\sigma_{i,j} = \hat r_{\sigma(i), \sigma(j)}
\end{equation}

@@

When we use a coupling method and the prior changes, we  can apply \eqref{eq:changePrior} either before or after coupling. If the results are identical, for all initial pairwise data $R$, then we say that the coupling method is Bayes covariant. Symbolically, let us denote by  $\pi$ the change of priors. Let $R^\pi$ be pairwise datum when we apply \eqref{eq:changePrior} to each binary classifier.
For Bayes covariant method $V$ the following diagram commutes.

\begin{figure}[!h]
\centering
\begin{tikzcd}
	R \arrow[r, "\pi"] \arrow[d, "V"] & R^\pi \arrow[d, "V"] \\
	V(R) \arrow[r, "\pi"] & V(R^\pi) = V(R)^\pi
\end{tikzcd}
\caption{Diagrammatical description of Bayes covariance for a coupling method $V$}
\label{fig:bc}
\end{figure}

@@

@@

Let us finally describe \emph{Hinton's minimality}. Geoffrey Hinton is reported to object to the underlying principle of one-vs-one classification, where any single classifier influences the multi-class prediction \cite[p.~467]{hastie1998classification}. Suppose we are classifying objects from CIFAR-10 dataset.  If the true class is  a ``dog'', why should we care what is the output of the classifier distinguishing ``airplanes'' from ``ships''? The binary classifier is trained only on airplanes and ships, and has never seen a dog. One may thus prefer a classifier which somehow suppresses noisy output of (seemingly) irrelevant classifiers.

The first that the five conditions have a clear mathematical definition, whereas the rest are not exact. For instance while the ideal situation for a coupling method is to have no parameters, the presence of one or two trainable parameters may be justified, if it yields noticeably better performance.

@@

There is a plethora of methods available for coupling. Let us describe four major categories.

\subsection{Optimizing binary divergence}

@@

Their method use an additional trick, which we term \emph{self-referentiality}. In the method of Hastie and Tibshirani, $w_{ij}$ are constants derived from observed class frequencies. However, we could use weights $w_{ij}$ that increase when the predicted probability of either class $i$ or $j$ is high. A reasonable choice would be for instance $w_{ij}= p_i + p_j$. In the case of quadratic score, it is more convenient to use weights $w_{ij} = (p_i + p_j)^2$ because then the resulting optimization problem is just minimization of a quadratic form
\begin{align*}
\hat{\boldsymbol{p}} = \argmin_{\boldsymbol{p}} \sum_{i,j} (r_{ij}p_j - r_{ji}p_i)^2.
\end{align*}

@@

In order to convert Hinton's oracle into a true classification method we can use the principle of self-referentiality.  Denote by $\hat{\boldsymbol{T}}_i$ the prediction of the arboreal method, where the underlying tree is a star centered at class $C_i$. The predictions will generally span the simplex $\sum_i \hat p_i = 1$. In that case the prediction $\hat{\boldsymbol{p}}$ of the method  can be expressed as a linear combination of $\hat{\boldsymbol{T}}_i$. Let us now define \emph{radial} coupling method. The  method requires that for  weights we could in fact use the components of $\boldsymbol{p}$ i.e.
\begin{equation}
	\hat {\boldsymbol{p}} = \sum_{i=1}^K p_i \hat{\boldsymbol{T}}_i. \label{eq:radial}
\end{equation}
%
The intuition behind \eqref{eq:radial} is that if some component of $\hat{\boldsymbol{p}}$ is close to 1 (and thus the remaining components are close to zero), then the prediction $\hat{\boldsymbol{p}}$ should be close to the prediction of Hinton's oracle which is provided by $\hat{\boldsymbol{T}}_i$.

@@

There are many Bayes covariant methods. For instance any arboreal method is Bayes covariant. Given two Bayes covariant methods one can interpolate between them using their ``linear combination'' - an ensemble which is also Bayes covariant. Let us make it explicit.

 Suppose real numbers $\alpha, \beta$ are given such $\alpha + \beta = 1$. If the prediction of the two Bayes covariant methods $\boldsymbol{B}$ and $\boldsymbol{\tilde B}$ are
\begin{align*}
	\boldsymbol{B}(x) = (p_1, \ldots, p_K)\quad\textrm{and}\quad \boldsymbol{\tilde B}(x) = (\tilde p_1, \ldots, \tilde p_K)
\end{align*}
then the prediction of the ensemble $\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}$ is given by
\begin{align*}
	\bigl(\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}\bigr)(x)\propto (\,p_1^\alpha \tilde p_1^\beta, \ldots, p_K^\alpha \tilde p_K^\beta).
\end{align*}

@@

>>>>

\newtheorem{prop}{Proposition}
\newtheorem{thm}{Theorem}
\newtheorem{assumption}{Assumption}

@@

\usepackage{lastpage}
\jmlrheading{23}{2025}{1-\pageref{LastPage}}{1/21; Revised 5/22}{9/22}{21-0000}{Author One and Author Two}

@@

\author{\name Ondrej Šuch \email ondrejs@savbb.sk \\
       \addr Matematický ústav SAV\\
       Ďumbierska 1\\
       Banská Bystrica, 974 01, Slovakia
       \AND
       \name Peter Novotný \email peter.novotny@uniza.sk \\
       \addr Fakulta informatiky a riadenia\\
       Žilinská Univerzita v Žiline\\
       Žilina, 010 26, Slovakia
       \AND
       \name Ali Haidar \email haidar@savbb.sk \\
       \addr Matematický ústav SAV\\
       Ďumbierska 1\\
       Banská Bystrica, 974 01, Slovakia
       }

@@

An exact solution to this problem is provided by the Bayes classifier. To simplify the discussion we will make the following assumption.

\begin{assumption} \label{ass:1}
 For all $x$ in the feature space we have $p_i(x) > 0$.
\end{assumption}
Commonly used classifiers such as linear discriminant analysis, (penalized) logistic regression or support vector machines all satisfy this assumption.

@@

\begin{align}
	p_{ij}^\Pi(x) = \frac{p_i(x) P_i}{p_i(x)P_i + p_j(x)P_j}.
\end{align}

Knowledge of vector $\boldsymbol{p}^\Pi(x)$ is  equivalent to knowing the functions $p^\Pi_{ij}(x)$. Indeed under Assumption \ref{ass:1} we have
\begin{align}
p_{ij}^\Pi(x) = \frac{p_i^\Pi(x)}{p_i^\Pi(x) + p_j^\Pi(x)}. \label{eq:bt1}
\end{align}

@@

In practice one does not have know true probability distributions and hence cannot construct Bayes classifier. However, given a sample from the distribution, we can train binary classifiers to provide approximations $r_{ij}(x)$ to pairwise Bayes classifiers $p_{ij}^\Pi(x)$. The goal of probabilistic one-vs-one classification is to deduce for any $x$ in the feature space $X$ an estimate $\hat{\boldsymbol{p}}(x)$ for $\boldsymbol{p}^\Pi(x)$. One does it by finding an approximate solution of the system of equations
%\begin{align}
%
%\end{align}
%
%
%Suppose that $K$ classes $C_1, \ldots, C_K$ are distributed according to probability distributions $p_i$ on a  space $X$. Let us denote the $r_{ij}(x)$ the output of Bayes (binary) classifier for $x$ in $X$ yielding the probability $p(x \in C_i \mid x\in C_i \cup C_j )$. Then we have
%\begin{equation}
%r_{ij}(x)= \frac {p_i(x)}{p_i(x) + p_j(x)}.
%\end{equation}
%
%
%The system of equations is called Bradley--Terry equations.  The equations in \eqref{eq:bt1}  can be transformed to a system of linear equations. The theoretical basis for one-vs-one classification framework is provided by the following result.
%
%
%
%Suppose we have estimators $\hat r_{ij}$ for Bayes predictions $r_{ij}$ for all pairs $i\not= j$. A \emph{coupling method} obtains a multi-class probabilistic estimate $\boldsymbol{\hat p}= (\hat p_1, \ldots, \hat p_K)$ that satisfy
\begin{equation}
	{r}_{ij}(x) \approx \frac {\hat p_i(x)}{\hat p_i(x) + \hat p_j(x)}. \label{eq:bt2}
\end{equation}

@@

Equations forming system \eqref{eq:bt2} are called Bradley-Terry equations.
The system is usually inconsistent and additional assumptions are needed. A method to solve Bradley-Terry equations is called a coupling method. Its input can be represented by a matrix $\boldsymbol{R}(x)$ with off-diagonal entries being $r_{ij}(x)$. Schematically, classification using coupling method $\boldsymbol{C}$ proceeds as indicated in Figure \ref{fig:coupling}.

\begin{figure}[!h]
	\centering
	\begin{tikzcd}[column sep=3cm]
		\textrm{sample $x$}\arrow[r, "binary~classifiers"] \arrow{dr}[swap]{multi-class~ prediction} & \boldsymbol{R}(x) \arrow{d}{coupling}[swap]{V} \\
		 & V(\boldsymbol{R}(x)) 
	\end{tikzcd}
	\caption{Multi-class classification using a coupling method ${V}$}
	\label{fig:coupling}
\end{figure}

@@

Let us outline what kind of conditions may be desirable in a coupling method. In %Section \ref{sec:des:exact} 
this section
 we will describe exact mathematical criteria, while in section \ref{sec:des:inexact} we shall describe more flexible properties.

@@

\emph{Symmetry} refers to the action of the permutation group which acts both on the classes as well as on multi-class predictions. The action of a permutation $\sigma$ extends to the action on the pairwise matrix $\boldsymbol{R} = (r_{ij})$ by 
\begin{equation}
	 r^\sigma_{i,j} \stackrel{def}{=}  r_{\sigma(i), \sigma(j)}.
\end{equation}
We say that a coupling method $V$ is \emph{symmetric} if 
\begin{align}
		V(\boldsymbol{R})^\sigma = V(\boldsymbol{R}^\sigma)\quad\textrm{for any permutation $\sigma$}.
\end{align}

@@

When we use a coupling method and the prior changes, we  can apply \eqref{eq:changePrior} either before or after coupling. If the results are identical, for all initial pairwise data $\boldsymbol{R}$, then we say that the coupling method is Bayes covariant. Symbolically, let us denote by  $\pi$ the change of priors. Let $\boldsymbol{R}^\pi$ be pairwise datum when we apply \eqref{eq:changePrior} to each binary classifier.
For Bayes covariant method $V$ the following diagram commutes.

\begin{figure}[!h]
\centering
\begin{tikzcd}
	\boldsymbol{R} \arrow[r, "\pi"] \arrow[d, "V"] & \boldsymbol{R}^\pi \arrow[d, "V"] \\
	V(\boldsymbol{R}) \arrow[r, "\pi"] & V(\boldsymbol{R}^\pi) = V(\boldsymbol{R})^\pi
\end{tikzcd}
\caption{Diagrammatical description of Bayes covariance for a coupling method $V$}
\label{fig:bc}
\end{figure}

@@

In this section we describe several concepts that lack strict mathematical delineation, but may still be important to consider.

@@

Let us finally mention \emph{Hinton's minimality}. Geoffrey Hinton is reported to object to the underlying principle of one-vs-one classification, where any single classifier influences the multi-class prediction \cite[p.~467]{hastie1998classification}. Suppose we are classifying objects from CIFAR-10 dataset.  If the true class is  a ``dog'', why should we care what is the output of the classifier distinguishing ``airplanes'' from ``ships''? The binary classifier is trained only on airplanes and ships, and has never seen a dog. One may thus prefer a classifier which somehow suppresses noisy output of (seemingly) irrelevant classifiers.

%The first that the five conditions have a clear mathematical definition, whereas the rest are not exact. For instance while the ideal situation for a coupling method is to have no parameters, the presence of one or two trainable parameters may be justified, if it yields noticeably better performance.

@@

There is a plethora of methods available for coupling. Let us describe four major classes.

\subsection{Methods aiming to minimize binary divergence}

@@

Their method uses an additional trick, which we term \emph{self-referentiality}. In the method of Hastie and Tibshirani, $w_{ij}$ are constants derived from observed class frequencies. However, we could use weights $w_{ij}$ that increase when the predicted probability of either class $i$ or $j$ is high. A reasonable choice would be for instance $w_{ij}= p_i + p_j$. In the case of quadratic score, it is more convenient to use weights $w_{ij} = (p_i + p_j)^2$ because then the resulting optimization problem is just minimization of a quadratic form
\begin{align*}
\hat{\boldsymbol{p}} \stackrel{def}{=} \argmin_{\boldsymbol{p}} \sum_{i,j} (r_{ij}p_j - r_{ji}p_i)^2.
\end{align*}

@@

In order to convert Hinton's oracle into a true classification method we can use the principle of self-referentiality.  Denote by $\hat{\boldsymbol{T}}_i$ the prediction of the arboreal method, where the underlying tree is a star centered at class $C_i$. The predictions will generally span the simplex $\sum_i \hat p_i = 1$. In that case the prediction $\hat{\boldsymbol{p}}$ of the method  can be expressed as a linear combination of $\hat{\boldsymbol{T}}_i$. Let us now define \emph{radial} coupling method. The  method requires that for  weights we could in fact use the components of $\boldsymbol{p}$ i.e.
\begin{equation}
	\hat {\boldsymbol{p}} = \bigoplus_{i=1}^K p_i \hat{\boldsymbol{T}}_i. \label{eq:radial}
\end{equation}
%
The intuition behind \eqref{eq:radial} is that if some component of $\hat{\boldsymbol{p}}$ is close to 1 (and thus the remaining components are close to zero), then the prediction $\hat{\boldsymbol{p}}$ should be close to the prediction of Hinton's oracle which is provided by $\hat{\boldsymbol{T}}_i$.

@@

There are many Bayes covariant methods. For instance any arboreal method is Bayes covariant.

Given several Bayes covariant methods one can form  their ``linear combination'' - an ensemble which is also Bayes covariant. Let us make it explicit.

 Suppose real numbers $\alpha_1, \alpha_L$ are real numbers and $\boldsymbol{C}_1, \ldots, \boldsymbol{C}_L$ are coupling methods. We define their linear combination by requiring that 
 \begin{align*}
 	\bigl(\alpha_1 \boldsymbol{C}_1 \oplus \ldots \oplus \alpha_L \boldsymbol{C}_L\bigr)(x) \propto 
 \end{align*}

 given such $\alpha + \beta = 1$. If the prediction of the two Bayes covariant methods $\boldsymbol{B}$ and $\boldsymbol{\tilde B}$ are
\begin{align*}
	\boldsymbol{B}(x) = (p_1, \ldots, p_K)\quad\textrm{and}\quad \boldsymbol{\tilde B}(x) = (\tilde p_1, \ldots, \tilde p_K)
\end{align*}
then the prediction of the ensemble $\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}$ is given by
\begin{align*}
	\bigl(\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}\bigr)(x)\propto (\,p_1^\alpha \tilde p_1^\beta, \ldots, p_K^\alpha \tilde p_K^\beta).
\end{align*}

@@

We considered the following questions

- are there Bayes covariant methods that match the performance of Wu-Lin-Weng's method ?
- when priors change, is it better to apply Bayes update before or after coupling?
- which coupling methods are best with respect to Hinton's minimality?

\subsection{Datasets}

\begin{figure}[!ht]
	\includegraphics[width = 0.5\linewidth]{graph/dna-lda.pdf}
	\includegraphics[width = 0.5\linewidth]{graph/waveform-lda.pdf}
	\caption{Projections of three-class datasets using LDA Left: dna, right: waveform}
\end{figure}
